---
title: "Predicting Human Activity Using Accelerometer Data"
author: "Degu Hewan Demissie"
date: "2025-09-21"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r package, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(caret)
library(randomForest)
library(dplyr)
library(ggplot2)
```


# Introduction

This project applies machine learning techniques to predict human activity based on accelerometer data collected from wearable devices. The goal is to build a robust classifier using the  variable as the outcome and evaluate its performance.

# Load data

```{r load-data}
training <- read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv")
testing  <- read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv")
# Remove columns with >90% missing values
training <- training[, colSums(is.na(training)) < nrow(training) * 0.9]
```

# Remove columns with too many NAs and irrelevant metadata

```{r remove unwanted-data}
training <- training %>%
  select_if(function(col) mean(!is.na(col)) > 0.9) %>%
  select(-c(X, user_name, raw_timestamp_part_1, raw_timestamp_part_2,
            cvtd_timestamp, new_window, num_window))
head(training)
```

# Partition the data

```{r partiction-data}
set.seed(123)
inTrain <- createDataPartition(training$classe, p = 0.7, list = FALSE)
trainSet <- training[inTrain, ]
validationSet <- training[-inTrain, ]

training <- training[, -c(1:7)]
```

# Ensure target variable is intact


```{r intact-data}
table(training$classe)
```

#  Minimal Working Example

```{r minimal-data}
library(caret)
library(randomForest)
```

# Ensure target is factor

```{r target-data}
# Preserve 'classe' before filtering
classe_col <- trainSet$classe

# Remove character columns except 'classe'
trainSet <- trainSet[, sapply(trainSet, function(x) !is.character(x))]

# Reattach 'classe' and convert to factor
trainSet$classe <- as.factor(classe_col)
table(trainSet$classe)
trainSet$classe <- as.factor(trainSet$classe)

```
# Train model

```{r train1-data}
library(caret)
library(randomForest)
classe_col <- trainSet$classe
trainSet <- trainSet[, sapply(trainSet, function(x) !is.character(x))]
trainSet$classe <- as.factor(classe_col)

```

# Missing values
If  contains s,  may fail silently or throw cryptic errors.ix:


```{r fix-data}
trainSet <- trainSet[, colSums(is.na(trainSet)) == 0]
trainSet <- trainSet[!is.na(trainSet$classe), ]
```


# Empty or corrupted 

If  has zero rows or columns after preprocessing, the model can’t train.
heck:

```{r empty corrupted-data}
dim(trainSet)
summary(trainSet$classe)
```

# Final evaluation 

```{r train1_data}
library(caret)
library(randomForest)

set.seed(123)

# Ensure 'classe' is preserved and predictors are clean
classe_col <- trainSet$classe
trainSet <- trainSet[, sapply(trainSet, function(x) !is.character(x))]
trainSet$classe <- as.factor(classe_col)

# Remove predictors with missing values
trainSet <- trainSet[, colSums(is.na(trainSet)) == 0]

model_rf <- train(classe ~ ., data = trainSet,
                  method = "rf",
                  trControl = trainControl(method = "cv", number = 5))
model_rf
```

